Computer-Assisted Analysis of the \\ Anderson-Hájek Ontological Controversy

### 2017-01-04 Papers/2017/LogicaUniversalis-HajekAndersonControversy/main.tex
Fix a minor conceptual mistake about the notion of contingency.
Fix minor typos and mistakes.

The particular motivation and context for the work presented in the given
article is as follows: It is well known that the axioms in G\"odel's
ontological proof (cf. Fig. \ref{fig:Goedel}) entail what is called
\emph{modal collapse} \citep{Sobel1987,SobelBook2004}. This means that
the formula $\varphi \rightarrow \Box \varphi$, abbreviated as MC,
holds for any formula $\varphi$ and not just for
$\exists x. \mathit{God}(x)$ as intended.  Hence, contingent truth
implies necessary truth, or, in other words, there are no unnecessary
contingent truths. In fact, one may even go as far and interpret MC as a result
against free will.
%
The fact that MC follows from Gödel's axioms, which has also been
confirmed with higher-order automated theorem provers in our previous
experiments \citep{C40,J30}, has led to strong criticism of the
argument and stimulated attempts to remedy the problem.
\citet{Hajek_Magari_and_others_1996,Hajek_der_Mathematiker_2001}
proposed the use of cautious instead of full comprehension principles,
and \citet{fitting02:_types_tableaus_god} took greater care of the
semantics of higher-order quantifiers in the presence of
modalities. Others, such as
\citet{anderson90:_some_emend_of_goedel_ontol_proof},
\citet{Hajek2002} and \citet{bjordal99}, proposed emendations of
G\"odel's axioms and definitions. They require neither comprehension
restrictions nor more complex semantics and are therefore technically
simpler to analyze with computer support. Hence, we here concentrate
on the latter emendations of Gödel's argument and leave the former
ones for future work.  Again, familiarity with the papers by
\citet{anderson90:_some_emend_of_goedel_ontol_proof},
\citet{Hajek2002} and \citet{bjordal99} is assumed, since we here are only
briefly commenting/assessing various claims made in there.

Our formalizations employ the embedding of higher-order modal logic
(HOML) in classical higher-order logic (HOL) as introduced in previous
work \citep{C40,J23} as enabling technology.  We explored the effect
of different domain conditions on the provability of lemmas, theorems
and even axioms.  This was motivated by a controversy between Hájek
and Anderson regarding the redundancy of some axioms in Anderson's
emendation. In \emph{constant domain semantics} (\emph{possibilist
  notion of quantification}), the individual domains are the same in
all possible worlds. In \emph{varying domain semantics}
(\emph{actualist notion of quantification}), the domains may vary from
world to world. The latter notion is technically encoded in our
approach with the help of an existence relation expressing which
individuals actually exist in each world. In other words, actualist
quantification is formalized as possibilist quantification guarded by
the existence relation.\footnote{EDNOTE: This part was
  misleading/wrong before; I reworded it.} This technical solution
enables us to flexibly switch between these different notions of
quantification and to even mix them. When we mention varying
domain semantics (actualist quantification) below, we in fact mean
varying domain semantics for the domain of individuals only, while for
the other types we still assume the standard constant domain semantics
(possibilist quantification).  This is, to our best knowledge, in line 
with the intention of the authors (it would of course be easily
possible in our approach to study other combinations).

@@

Our main results, which we have
presented at the 1st World Congress on Logic and Religion \citep{C41},
are presented in more detail below. The corresponding formalizations,
containing all the details, are available online: see the
subdirectories \url{Anderson}, \url{Hajek} and \url{Bjordal} at
\url{github.com/FormalTheology/GoedelGod/blob/master/Formalizations/Isabelle/}.
\footnote{Ideally, Logica Universalis should provide means for
  permanent storing of such additional material.}

@@

\begin{figure}
\begin{sideways}
%\includegraphics[width=1.2\textwidth]{Summary.png} \fbox{EDNOTE: to be removed}
\end{sideways}
%
\begin{sideways}
%\scalebox{.9}{
\resizebox{19cm}{4.5cm}{
\begin{tabular}{lccccccccccccc} 
\toprule
Investigated Variant & D1$'$ & A:A1$'$ & A2$'$ & A3$'$ & A4 & A4$'$ & H:A4 & A5 & A5$'$
  & H:A5 & T3 & T3$'$ & MC \\  
\midrule 
Scott (const) & - & - & - & - & \tabboxs{\tabred}{N/I} & - & - & \tabboxs{\tabred}{N/I} & - & - & P & -  & P \\
Scott (var) & - & - & - & - & \tabboxs{\tabred}{N/I} & - & - & \tabboxs{\tabred}{N/I} & - & - & P & -  & P \\
Anderson (const) & - & - & - & - & \tabbox{\tabred}{R(K4B)} & - & - & - & \tabboxs{\tabred}{R}  & - & - & P  & CS \\
Anderson (var) & - & - & - & - & \tabbox{\tabred}{R(K4B)} & - & - & - & \tabboxs{\tabred}{R}  & - & - & P  & CS \\
Anderson (mix) & - & - & - & - & \tabbox{\tabred}{R(K4B)} & - & - & - & \tabboxs{\tabred}{I}  & - & - & \tabboxs{\tabred}{CS} & CS \\
H\'ajek AOE$'$ (var) & - & - & \tabboxs{\tabyellow}{CS}  &  & \tabboxs{\tabred}{S/I} & - & - & - & \tabboxs{\tabred}{S/I}  & - & - & P(KB) & CS \\
H\'ajek AOE$'$\_0 (var) & - & - & - & \tabboxs{\tabyellow}{CS} & \tabboxs{\tabgreen}{R} & - & - & - & S/U & - & - & P(KB) & CS \\
H\'ajek AOE$''$ (var) & - & - &  &  & - & - & \tabboxs{\tabred}{S/I} & - & - & \tabboxs{\tabred}{S/I} & - & P(KB) & CS \\
Anderson (simp) (var) & - & R & R  &  &  & R(K4B) & - & - & & - & - &  & \\
Bjørdal (const) & \tabbox{\tabgreen}{R(K4)} & - & \tabboxs{\tabgreen}{R} & \tabboxs{\tabgreen}{R} & - & \tabbox{\tabgreen}{R(KT)} & - & - &\tabboxs{\tabyellow}{N/I} & - & - & \tabbox{\tabgreen}{P(KB)} &  CS \\
Bjørdal (var) & \tabboxs{\tabyellow}{CS} & - & \tabboxs{\tabgreen}{R} & \tabboxs{\tabgreen}{R} & - & \tabbox{\tabgreen}{R(KT)} & - & - &\tabboxs{\tabyellow}{N/I} & - & - & \tabbox{\tabgreen}{P(KB)} &  CS \\
\bottomrule
\end{tabular}
}
\end{sideways}
\caption{Summary of Results.}
\label{fig:summary}
\end{figure}

@@

\section{Conclusion}
Using our approach, that is, semantical embeddings of variants of higher-order
modal logics in classical higher-order logic, the formalization and
(partly) automated analysis of several variants of G\"odel's
ontological argument has been surprisingly straightforward. The
higher-order provers we employed not only confirmed many claimed results, but also exposed a
few mistakes and novel insights in a long standing controversy.  We
believe the technology employed in this work is ready to be fruitfully
adopted in larger scale by philosophers.  Moreover, our approach is
by no means restricted to metaphysics or philosophy, and further work
includes similar applications in other areas, including, for example,
law, politics and ethics. In fact, we claim, that our approach, at
least to some degree, a realises Leibniz' dream of a characteristica universalis
and calculus ratiocinator.

>>>>

The particular motivation and context for the work presented in the given
article is as follows: It is well known that the axioms in G\"odel's
ontological proof (cf. Fig. \ref{fig:Goedel}) entail what is called
\emph{modal collapse} \citep{Sobel1987,SobelBook2004}. This means that
the formula $\varphi \rightarrow \Box \varphi$, abbreviated as MC,
holds for any formula $\varphi$ and not just for
$\exists x. \mathit{God}(x)$ as intended.  Hence, anything that is the case is necessarily the case, actual truth implies necessary truth, or, in other words, there are no
contingent truths. In fact, one may even go further and interpret MC as a result
against free will.
%
The fact that MC follows from Gödel's axioms, which has also been
confirmed with higher-order automated theorem provers in our previous
experiments \citep{C40,J30}, has led to strong criticism of the
argument and stimulated attempts to remedy the problem.
\citet{Hajek_Magari_and_others_1996,Hajek_der_Mathematiker_2001}
proposed the use of cautious instead of full comprehension principles,
and \citet{fitting02:_types_tableaus_god} took greater care of the
semantics of higher-order quantifiers in the presence of
modalities. Others, such as
\citet{anderson90:_some_emend_of_goedel_ontol_proof},
\citet{Hajek2002} and \citet{bjordal99}, proposed emendations of
G\"odel's axioms and definitions. They require neither comprehension
restrictions nor more complex semantics and are therefore technically
simpler to analyze with computer support. Hence, we here concentrate
on the latter emendations of Gödel's argument and leave the former
ones for future work.  Again, familiarity with the papers by
\citet{anderson90:_some_emend_of_goedel_ontol_proof},
\citet{Hajek2002} and \citet{bjordal99} is assumed, since we here are only
briefly commenting/assessing various claims made in there.

Our formalizations employ the embedding of higher-order modal logic
(HOML) in classical higher-order logic (HOL) as introduced in previous
work \citep{C40,J23} as enabling technology.  We explored the effect
of different domain conditions on the provability of lemmas, theorems
and even axioms.  This was motivated by a controversy between Hájek
and Anderson regarding the redundancy of some axioms in Anderson's
emendation. In \emph{constant domain semantics} (\emph{possibilist
  notion of quantification}), the individual domains are the same in
all possible worlds. In \emph{varying domain semantics}
(\emph{actualist notion of quantification}), the domains may vary from
world to world. The latter notion is technically encoded in our
approach with the help of an existence relation expressing which
individuals actually exist in each world. In other words, actualist
quantification is formalized as possibilist quantification guarded by
the existence relation. This technical solution
enables us to flexibly switch between these different notions of
quantification and to even mix them. When we mention varying
domain semantics (actualist quantification) below, we in fact mean
varying domain semantics for the domain of individuals only, while for
the other types we still assume the standard constant domain semantics
(possibilist quantification).  This is, to our best knowledge, in line 
with the intention of the authors (it would of course be easily
possible in our approach to study other combinations).

@@

Our main results, which we have
presented at the 1st World Congress on Logic and Religion \citep{C41},
are presented in more detail below. The corresponding formalizations,
containing all the details, are available online: see the
subdirectories \url{Anderson}, \url{Hajek} and \url{Bjordal} at
\url{github.com/FormalTheology/GoedelGod/blob/master/Formalizations/Isabelle/}.
\footnote{Ideally, Logica Universalis should provide
  permanent storage of such additional material.}

@@

\begin{figure}
\begin{sideways}
%\includegraphics[width=1.2\textwidth]{Summary.png} 
\end{sideways}
%
\begin{sideways}
%\scalebox{.9}{
\resizebox{19cm}{4.5cm}{
\begin{tabular}{lccccccccccccc} 
\toprule
Investigated Variant & D1$'$ & A:A1$'$ & A2$'$ & A3$'$ & A4 & A4$'$ & H:A4 & A5 & A5$'$
  & H:A5 & T3 & T3$'$ & MC \\  
\midrule 
Scott (const) & - & - & - & - & \tabboxs{\tabred}{N/I} & - & - & \tabboxs{\tabred}{N/I} & - & - & P & -  & P \\
Scott (var) & - & - & - & - & \tabboxs{\tabred}{N/I} & - & - & \tabboxs{\tabred}{N/I} & - & - & P & -  & P \\
Anderson (const) & - & - & - & - & \tabbox{\tabred}{R(K4B)} & - & - & - & \tabboxs{\tabred}{R}  & - & - & P  & CS \\
Anderson (var) & - & - & - & - & \tabbox{\tabred}{R(K4B)} & - & - & - & \tabboxs{\tabred}{R}  & - & - & P  & CS \\
Anderson (mix) & - & - & - & - & \tabbox{\tabred}{R(K4B)} & - & - & - & \tabboxs{\tabred}{I}  & - & - & \tabboxs{\tabred}{CS} & CS \\
H\'ajek AOE$'$ (var) & - & - & \tabboxs{\tabyellow}{CS}  &  & \tabboxs{\tabred}{S/I} & - & - & - & \tabboxs{\tabred}{S/I}  & - & - & P(KB) & CS \\
H\'ajek AOE$'$\_0 (var) & - & - & - & \tabboxs{\tabyellow}{CS} & \tabboxs{\tabgreen}{R} & - & - & - & S/U & - & - & P(KB) & CS \\
H\'ajek AOE$''$ (var) & - & - &  &  & - & - & \tabboxs{\tabred}{S/I} & - & - & \tabboxs{\tabred}{S/I} & - & P(KB) & CS \\
Anderson (simp) (var) & - & R & R  &  &  & R(K4B) & - & - & & - & - &  & \\
Bjørdal (const) & \tabbox{\tabgreen}{R(K4)} & - & \tabboxs{\tabgreen}{R} & \tabboxs{\tabgreen}{R} & - & \tabbox{\tabgreen}{R(KT)} & - & - &\tabboxs{\tabyellow}{N/I} & - & - & \tabbox{\tabgreen}{P(KB)} &  CS \\
Bjørdal (var) & \tabboxs{\tabyellow}{CS} & - & \tabboxs{\tabgreen}{R} & \tabboxs{\tabgreen}{R} & - & \tabbox{\tabgreen}{R(KT)} & - & - &\tabboxs{\tabyellow}{N/I} & - & - & \tabbox{\tabgreen}{P(KB)} &  CS \\
\bottomrule
\end{tabular}
}
\end{sideways}
\caption{Summary of Results.}
\label{fig:summary}
\end{figure}

@@

\section{Conclusion}
Using our approach, that is, semantical embeddings of variants of higher-order
modal logics in classical higher-order logic, the formalization and
(partly) automated analysis of several variants of G\"odel's
ontological argument has been surprisingly straightforward. The
higher-order provers we employed not only confirmed many claimed results, but also exposed a
few mistakes and novel insights in a long standing controversy.  We
believe the technology employed in this work is ready to be fruitfully
adopted in larger scale by philosophers.  Moreover, our approach is
by no means restricted to metaphysics or philosophy, and further work
includes similar applications in other areas, including, for example,
law, politics and ethics. In fact, we claim, that our approach, at
least to some degree, realizes Leibniz' dream of a characteristica universalis
and calculus ratiocinator.
